---
title: "bootcamp"
author: "Matt Pecsok"
date: "2023-08-21"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

The content in this bootcamp is covered by a huge number of web resources. Following up on your learning is easy and should be done soon and regularly to ensure you get the most value.

https://www.w3schools.com/r/default.asp
https://r4ds.had.co.nz/



```{r}
library(tidyverse)
```

# Basic Syntax and Variables

```{r}
print("hello world!")
```

```{r}
'Hello World!'
```

## Errors!

It's the best place to start. Because you'll create lots of them.

The first type of error we'll cover is syntax error. Get used to it.
Incorrect code will often generate syntax errors.

```{r}
#1 = 2
```

In programming, an expression is a combination of values, variables,
operators, and function calls that can be evaluated to produce a result.
Expressions are fundamental building blocks used to perform
calculations, manipulate data, and make decisions within a program. They
can represent a wide range of operations, from simple arithmetic
calculations to complex logical evaluations.

An expression is incomplete if it is missing something. In this case "

```{r}
#"Hello World
```

## Literals

character string literals

a string can contain, letters, numbers special characters etc.

```{r}
#"hello world!"
```

```{r}
#print("hello world!")
```

```{r}
print(paste('hello','world!'))
```

YOU DO. Create a code block, put a string literal it it and run the code
block.

## Numeric Literals

```{r}
4
```

```{r}
# YOU TRY pick a number. Any number. 
```

### Addition

```{r}
1+1
```

```{r}
# YOU TRY. Add two numbers. 
```

```{r}
# YOU TRY multiply two numbers
```


```{r}
1/3
```

```{r}
9L
```

```{r}
class(9L)
```

### Logical Literal

```{r}
if (FALSE) {
  print("it's true")
} else {
  print("it's false")
}
```

```{r}
if ((TRUE) && (TRUE)) {
  print("it's true")
} else {
  print("it's false")
}
```

```{r}
if ((TRUE) || (TRUE)) {
  print("it's true")
} else {
  print("it's false")
}
```
# Vector "literal"

```{r}
c(TRUE, FALSE, TRUE) | c(FALSE, TRUE, TRUE)
```

### Comments

```{r}
# This is a COMMENT

1+1 # this is a comment at the end of the line. 

#This is an error!
```
# Operators

Operate on variables and values. Operators perform operations.



### Arithmetic
+ * - / ^

### Assignment

<- 

### Comparison

==  - Is Equal?
!=  - Is Equal?
> - Is Greater than?
<
>= - Is greater than or Equal?
<=

Logical
&&  - AND
|| - OR
! - NOT

```{r}

```




# Variables

Your new best friend.

Variables store literals with a name for use later.

use <- to assign values (= works as well for most situations)

## Character strings

```{r}

my_name <- "Matt"

my_name
```

we use the class function and pass in my_name. the return value tells us
what type of variable my_name is. given that we assigned a string
literal to it, it should be no surprise that it is a character type
(character).

```{r}
class(my_name)
```

```{r}
# create a string variable and display it

first_name <- "Matthew"
```

Find the length of the character variable

```{r}
nchar(first_name)
```

```{r}
first_name[1:4] # Doesn't work the same as python
```

```{r}
substr(first_name, start = 1, stop = 4)
```

You try

```{r}
# create a new variable of your name
# get the length of your name
```

```{r}
# get the first 3 characters of your name
```

```{r}
# get the 3rd to last characters of your name
```

```{r}
grepl("Matt","Matthew")
```

## Numeric Variables

```{r}
# use grepl to see if Cat is in Catherine
```

```{r}
my_int <- 42
my_int
```

```{r}
class(my_int)
```

```{r}
class(42.5)
```

You try

```{r}
# create a numeric variable
```

```{r}
# check the class
```

```{r}
# add 5 to the variable, show the result. 
```

Boolean variables in R are known as logical

```{r}
my_bool <- TRUE

class(my_bool)
```

```{r}
5 < 4
```

```{r}
4 != 2
```

```{r}
five_lt_three <- 5 < 3

five_lt_three
```

You try create a new variable that has a boolean value based on two
numbers.

```{r}
# you put code here
```

```{r}
"matt" == "MATT"

```

```{r}
tolower("matt") == tolower("MATT")
```

## Vector Variables

```{r}
odd_nums <- c(1,3,5,7,9)

odd_nums
```

```{r}
length(odd_nums)
```

```{r}
sum(odd_nums)
```

You try

```{r}
 # create a numeric vector
```

```{r}
 # find length
```

```{r}
# get class
```

```{r}
# sum
```

```{r}
 # create a logical vector
```

```{r}
 # find length
```

```{r}
# get class
```

```{r}
# sum
```

```{r}
#set.seed(10)
my_numbers <- rnorm(10, mean = 10, sd = 5)
my_numbers
```


```{r}
mean(my_numbers)
```

```{r}
sd(my_numbers)
```



# if

```{r}

if (1==2) {
  print('it evaluated to true')
}
```



# if/else

```{r}
if (FALSE) {
  print("FALSE IS TRUE")
} else {
  print("FALSE IS FALSE")
}
```

You try

```{r}
 # create an if else that evaluates if 4 is less than 3. 

#if () {
#  print('4 is less than 3')
#} else {
#  print('4 is NOT less than 3')
#}
```

```{r}
my_name
```
```{r}
grepl("c","abdef")
```

```{r}
if (grepl("A", my_name)) {
  print("True: 'A' is in my_name")
} else {
  print("False: 'A' is not in my_name")
}
```
You try. create an if/else statement that tests if the price below is greater than or equal to 60. 

if it is, multiply price by .9 to give a 10% discount. 
otherwise multiply price bby 1.1 to increase the cost.
at the end (outside of the if/else) show the new price

```{r}
# YOU TRY

price <- 60


```





### else if

```{r}
# Example variables
x <- 10

# Check various conditions
if (x > 15) {
  print("x is greater than 15")
} else if (x > 5) {
  print("x is greater than 5 but not greater than 15")
} else {
  print("x is not greater than 5")
}
```
```{r}
# Example variables
x <- 100

# Check various conditions
if (x > 15) {
  print("x is greater than 15")
} else if (x > 5) {
  print("x is greater than 5 but not greater than 15")
} else {
  print("x is not greater than 5")
}
```





## Logical Operators

### AND

```{r}
# Example variables
x <- 10
y <- 8

# Check if both x is greater than 5 and y is greater than 7
if (x > 5 && y > 7) {
  print("Both conditions are TRUE")
} else {
  print("At least one condition is FALSE")
}
```

You try

write an if/else statement to see if the string is greater than 6 characters, and if it contains "c" then print('length gt 16 and contains c') else print('false either not gt 16 or doesn't contain c')

```{r}

some_string_var <- "Hi my name is Matthew and I'm a professor at the U. Nice to meet you!"
```

This is an opportunity to talk about ' and "

### OR

```{r}
# Example variables
x <- 3
y <- 8

# Check if either x is greater than 5 or y is greater than 7
if (x > 5 || y > 7) {
  print("At least one condition is TRUE")
} else {
  print("Both conditions are FALSE")
}
```

You try. Change this from || to &&
Also change the text to correctly say what happened in the evaluation

```{r}
# Example variables
x <- 3
y <- 8

# Check if either x is greater than 5 or y is greater than 7
if (x > 5 || y > 7) {
  #print("At least one condition is TRUE")
} else {
  #print("Both conditions are FALSE")
}
```


# Precedence

R follows common mathematics conventions

PEMDAS

Parentheses Exponents (Powers and Roots) Multiplication and Division
(from left to right) Addition and Subtraction (from left to right) if we
simply evaluated from left to right we might get the 2+3 = 5 and then 5
\* 4 = 20. This is not how we would expect evaluation to work in math
terms.

```{r}
2 - 3 * 4
```

```{r}
(2 - 3) * 4
```





# Functions

## simple hardcoded function

```{r}

return_three <- function() {
  return(3)
} 

return_three()
```

## function with a parameter

```{r}

add_three <- function(the_num) {
  return(the_num + 3)
} 

add_three(7)
```

```{r}

# you try. create a function that multiplies a number by 8 run the function to see the result
```

## function with 2 parameters

```{r}

add_two_numbers <- function(num1,num2) {
  return(num1 + num2)
} 

add_two_numbers(7,8)
```

```{r}

# you try. create a function that multiplies two numbers passed in as arguments. run the function to see the result
```

## function with parameters and default arguments

```{r}

add_two_numbers_w_default <- function(num1=3,num2=5) {
  return(num1 + num2)
} 

add_two_numbers_w_default()
```

```{r}

# you try. create a function that has argument defaults which multiplies two numbers passed in as arguments. run the function with and without passing arguments to see the result
```

The following function from stringr removes the letter "a" from the
character string "Matt"

Embed this into a new function called "remove_a" and call it with:
Alexander

```{r}
str_replace_all("Matt","a","")
```

```{r}

```

# DataFrames

## dataframe from vectors

In R, we don't need a special library (like Pandas) to create
dataframes.

```{r}
names <- c('Jose','Selena','Jose','John','Hannah')
names[2]
```

```{r}
ages <- c(23,17,20,25,22)
ages[2]
```

```{r}
person_df = data.frame(names,ages)
person_df
```

```{r}
str(person_df)
```

```{r}
summary(person_df)
```

You try. create 2 vectors, one numeric with 3 prices. one character with
3 product names, apples, oranges, lemons

Create a dataframe with these 2 vectors. show the dataframe show the
structure and summary of the dataframes.

```{r}
# create the 2 vectors here
```

```{r}
# create the dataframe here. give it a name. like fruit. show the dataframe
```

```{r}
# str()
```

```{r}
# summary()
```

While base R has dataframes. Tidyverse is much better for working with
dataframes.

## tidyverse

### piping data

pipe sends the data to the first argument of the function.

here's the base R way of calling a function with an argument.

```{r}
nchar("Matt")
```

here is the tidyverse way to do the same thing.

%>% is known as piping the data.

```{r}
"Matt" %>% nchar()
```

```{r}

c(1,2,3,4) %>% length()
  
```

```{r}
person_df %>% filter(ages > 20)
```

```{r}
person_df %>% filter(ages <= 20)
```

```{r}
# You try. Filter the dataframe for ages greater than or equal to 18
```

```{r}
person_df['ages'] > 20
```

```{r}
person_df$gt_20 <- person_df$ages > 20

person_df %>% head()
```

```{r}
person_df %>% filter(names %in% c('Jose','Selena'))
```

```{r}
 # you try. filter for the name Selena or Hannah
```

```{r}
person_df %>% select(names,gt_20) %>% head(2)
```

## titanic dataframe/tibble

A tibble is a dataframe, plus a lot more. We often say dataframe to keep
things simple, but when we read data in with tidyverse a tibble is
different than a base R dataframe.

```{r}

titanic <- read_csv('https://raw.githubusercontent.com/matthewpecsok/4482_fall_2022/main/data/titanic.train.csv')


```

```{r}
person_df %>% class()
```

```{r}
titanic %>% class()
```

```{r}
titanic %>% summary()
```

```{r}
titanic %>% head()
```

```{r}
titanic %>% head(n=9)
```

```{r}
titanic %>% summary()
```

```{r}
titanic %>% tally()
```

```{r}
titanic %>% group_by(Sex) %>% tally()
```

```{r}
titanic %>% mutate(my_new_column = "Hi!!!!") 
```

```{r}
# You try

# Mutate a new columns with the words "Hello World!". The column name can be "howdy"

```

## Aggregation and summarization

```{r}
titanic %>% 
  group_by(Survived,Sex) %>% 
  tally() %>% 
  group_by(Sex) %>% 
  mutate(proportion_survived = n/sum(n)) %>%
  arrange(Sex)
```

```{r}
titanic %>%
  group_by(Sex) %>%
  summarise(mean_value = mean(Age, na.rm = TRUE),
            sd_value = sd(Age, na.rm = TRUE))
```

```{r}
titanic %>%
  group_by(Pclass) %>%
  summarise(mean_value = mean(Fare, na.rm = TRUE),
            sd_value = sd(Fare, na.rm = TRUE))
```

```{r}
titanic$Pclass <- as.character(titanic$Pclass)
```

```{r}
titanic_mean_sd <- titanic %>%
  group_by(Survived) %>%
  summarise(mean_value = mean(Fare, na.rm = TRUE),
            sd_value = sd(Fare, na.rm = TRUE))


titanic_mean_sd
```

## GG Plots

```{r}
ggplot(data = titanic) + geom_bar(aes(x = Sex))

# can be rewritten as

titanic %>%
  ggplot() +
  geom_bar(aes(x = Sex))
```

You try

```{r}
# create a barplot of Survived
```

```{r}
ggplot(data = titanic, aes(x = Pclass, y = Fare)) +
  geom_boxplot() +
  labs(x = "Pclass", y = "Fare", title = "Titanic: Pclass vs. Fare")

# alternative syntax

titanic %>% ggplot() +
  geom_boxplot(aes(x = Pclass, y = Fare)) +
  labs(x = "Pclass", y = "Fare", title = "Titanic: Pclass vs. Fare")
```

You try.

```{r}
# create a boxplot of Pclass and Age
```




```{r}
ggplot(data = titanic, aes(x = Age, y = Fare)) +
  geom_point() +
  labs(x = "Age", y = "Fare", title = "Titanic: Age vs. Fare")
```

## Checking for nulls

```{r}
titanic %>%
  summarise(across(everything(), ~ sum(is.na(.)))) 
```

```{r}
# Count nulls in each column
null_counts <- colSums(is.na(titanic))

# Print the null counts
print(null_counts)
```

# For loops

```{r}

for (i in c(1:10)) {
  print(paste("we're on iteration",i))
}
```

```{r}
sum <- 0

for (i in c(1:10)) {
  sum <- sum + i
}

print(sum)
```
Don't write for loops unless you have to Vector notation is often more efficient and concise. 

```{r}
sum(c(1:10))
```

You try. This 

